---
title: "Homework 4"
author: "Ivan Wolansky, iaw2110"
date: "April 23, 2019"
output:
  pdf_document: default
  html_document: default
---

```{r}
set.seed(0)
```

Part 1: Inverse Transform Method

Consider the Cauchy random variable X with probability density function:

$f_X(x) = \dfrac{1}{\pi}\dfrac{1}{(1+x^{2})}, -\infty<x<\infty$

1. Let U be a uniform random variable over [0,1]. Find a transformation of U that allows you to simulate X from U.

1. (Solution)

$\int_0^x f(t)dt = \int_0^x \dfrac{1}{\pi(1+t^{2})}dt = \dfrac{1}{\pi}arctan(x) + \dfrac{1}{2}$

Therefore, $U = \dfrac{1}{\pi}arctan(x) + \dfrac{1}{2} \rightarrow X = tan(\pi(U-\dfrac{1}{2}))$

2. Write a R function called cauchy.sim that generates n simulated Cauchy random variables. The function should have the single input n and should use the inverse-transformation from Part 1. Test your function using 10 draws.

```{r}
cauchy.sim <- function(n) {
  u <- runif(n)
  return(ifelse((u<0|u>1), 0, tan(pi * (u-1/2))))
}
cauchy.sim(10)
```

3. Using your function cauchy.sim, simulate 1000 random draws from a Cauchy distribution. Store the 1000 draws in the vector cauchy.draws. Construct a histogram of the simulated Cauchy random variable with $f_X(x)$ overlaid on the graph. Note: when plotting the density curve over the histogram, include the argument prob = T. Also
note: the Cauchy distribution produces extreme outliers. I recommend plotting the histogram over the interval (-10, 10).

```{r}
cauchy.draws <- cauchy.sim(1000)
hist(cauchy.draws, prob = TRUE, breaks = 10000, xlim=c(-10, 10), ylim=c(0,0.35), 
     main="1000 Cauchy Draws")
x <- seq(-10, 10, .01)
lines(x, 1/(pi*(1+x^2)), col="blue")
```

Part 2: Reject-Accept Method

Problem 2

Let random variable X denote the temperature at which a certain chemical reaction takes
place. Suppose that X has probability density function:

$$f(x) = \begin{cases} 
          \frac{1}{9}(4-x^{2}), -1\leq x\leq 2 \\
          0, otherwise
        \end{cases}$$
Perform the following tasks:

4. Write a function f that takes as input a vector x and returns a vector of f(x) values.
Plot the function between -3 and 3. Make sure your plot is labeled appropriately

```{r}
f <- function(x) {
  return(ifelse((x < -1 | x > 2), 0, (1/9)*(4 - x^2)))
}
x <- seq(-3, 3, by = 0.01)
plot(x, f(x), main="Target Distribution", ylab="f(x)", type = "l", col="purple")

```

5. Determine the maximum of f(x) and find an envelope function e(x) by using a uniform
density for g(x). Write a function e which takes as input a vector x and returns a vector
of e(x) values.

The maximum of f(x) is 4/9 because the derivative of $\frac{1}{9}(4-x^2)$ is $-\frac{2x}{9}$, so the maximum is at x = 0, and f(0) = 4/9.

```{r}
f.max <- 4/9
e <- function(x) {
  return(ifelse((x < -1 | x > 2), Inf, f.max))
}
plot(x, f(x), main="Target Distribution", xlab="x", ylab="f(x)", type = "l", col="green")
lines(x, e(x), col="blue")
```

6. Using the Accept-Reject Algorithm, write a program that simulates 10,000 draws
from the probability density function f(x) from Equation 1. Store your draws in the
vector f.draws.

```{r}
samples <- 10000
n <- 0
f.draws <- numeric(samples)
while (n < samples) {
  y <- runif(1, min = -1, max=2)
  u <- runif(1)
  if (u < f(y)/e(y)) {
    n <- n + 1
    f.draws[n] <- y
  }
}
```

7. Plot a histogram of your simulated data with the density function f overlaid in the
graph. Label your plot appropriately.

```{r}
hist(f.draws, prob = TRUE, ylab="f(x)", xlab="x", 
     main="Histogram of Reject-Accept Draws from Equation 1", xlim = c(-3, 3))
lines(x, f(x), col="red")
```

Problem 3: Reject-Accept Method Continued

Consider the standard normal distribution X with probability density function:

$f(x) = \dfrac{1}{\sqrt{2\pi}}$exp$(-\dfrac{x^{2}}{2}), -\infty < x < \infty$)

In this exercise, we will write a function named normal.sim that simulates a standard
normal random variable using the Accept-Reject Algorithm.

Perform the following tasks:

8.  Write a function f that takes as input a vector x and returns a vector of f(x) values.
Plot the function between -5 and 5. Make sure your plot is labeled appropriately.

```{r}
f <- function(x) {
  return(1/sqrt(2*pi)*exp(-x^2/2))
}
x <- seq(-5, 5, by = 0.01)
plot(x, f(x), main="Standard Normal", type="l")
```


9. Let the known density g be the Cauchy density defined by pdf:

$g(x) = \dfrac{1}{\pi(1=x^2)}, -\infty < x < \infty$


Write a function e that takes as input a vector x and constant alpha (0 < $\alpha$ < 1)
and returns a vector of e(x) values. The envelope function should be defined as
e(x) = g(x)/$\alpha$.

```{r}
e <- function(x, alpha=1) {
  return((1/(pi*(1+x^2))) / alpha)
}
```


10. Determine a "good" value of $\alpha$. You can solve this problem graphically. To show your
solution, plot both f(x) and e(x) on the interval [-10, 10].

```{r}
x <- seq(-10, 10, by = 0.01)
alpha <- seq(0, 0.75, by = 0.15)
plot(x, f(x), main="Cauchy Envelope On Standard Normal", type="l", ylim=c(0,0.60))
for (i in alpha) {
  lines(x, e(x, alpha=i), col="red")
}
```
From looking at the graph, a good alpha value is somewhere around alpha = 0.6.

11. Write a function named normal.sim that simulates n standard normal random variables using the Accept-Reject Algorithm. The function should also use the Inverse Transformation from Part 1. Test your function using n=10 draws.

```{r}
normal.sim <- function(samples) {
  n <- 0
  normal.draws <- numeric(samples)
  while (n < samples) {
    y <- cauchy.sim(1)
    u <- runif(1)
    if (u < f(y)/e(y)) {
      n <- n + 1
      normal.draws[n] <- y
    }
  }
  return(normal.draws)
}
normal.sim(10)
```

12. Using your function normal.sim, simulate 10,000 random draws from a standard
normal distribution. Store the 10,000 draws in the vector normal.draws. Construct a histogram of the simulated standard normal random variable with f(x) overlaid on the graph. Note: when plotting the density curve over the histogram, include the
argument prob = T.

```{r}
normal.draws <- normal.sim(10000)
hist(normal.draws, prob = TRUE, ylab="f(x)", xlab="x", 
     main="Histogram of Reject-Accept Draws", xlim = c(-5, 5), breaks = 40)
lines(x, f(x), col="red")
```
Part 3: Simulation with Built-in R Functions
\begin{itemize}
  \item Start with x = 5
  \item Draw a random number r uniformly between -2 and 1
  \item Replace x with x + r
  \item Stop if x $\leq$0
  \item Else repeat
\end{itemize}

Perform the following tasks:

13. Write a while() loop to implement this procedure. Importantly, save all the positive
values of x that were visited in this procedure in a vector called x.vals, and display
its entries.

```{r}
x.vals <- 5
i <- 1
while (x.vals[i] > 0) {
  r <- runif(1, min = -2, max = 1)
  temp <- x.vals[i]
  if (temp + r < 0) {
    break
  } else {
    x.vals <- c(x.vals, x.vals[i] + r)
  }
  i <- i + 1
}
x.vals
```

14. Produce a plot of the random walk values x.vals from above versus the iteration
number. Make sure the plot has an appropriately labeled x-axis and and y-axis. Also
use type="o" so that we can see both points and lines.

```{r}
plot(1:length(x.vals), x.vals, type = "o",
         xlab="Iteration", ylab="Random Walk")
```

15. Write a function random.walk() to perform the random walk procedure that you
implemented in question (9). Its inputs should be: x.start, a numeric value at which
we will start the random walk, which takes a default value of 5; and plot.walk, a
boolean value, indicating whether or not we want to produce a plot of the random
walk values x.vals versus the iteration number as a side effect, which takes a default
value of TRUE. The output of your function should be a list with elements: x.vals,
a vector of the random walk values as computed above; and num.steps, the number
of steps taken by the random walk before terminating. Run your function twice with
the default inputs, and then twice times with x.start equal to 10 and plot.walk =
FALSE.

```{r}
random.walk <- function(x.start= 5, plot.walk = TRUE){
  x.vals <- 5
  i <- 1
  while (x.vals[i] > 0) {
    r <- runif(1, min = -2, max = 1)
    temp <- x.vals[i]
    if (temp + r < 0) {
      break
    }else {
      x.vals <- c(x.vals, x.vals[i] + r)
    }
    i <- i + 1
  }
  if (plot.walk) {
    plot(1:length(x.vals), x.vals, type = "o",
         xlab="Iteration", ylab="Random Walk")
  }
  return(list(x.vals = x.vals, num.steps = i))
}
random.walk()
random.walk()
random.walk(10, FALSE)
random.walk(10, FALSE)
```

16. We'd like to answer the following question using simulation: if we start our random
walk process, as defined above, at x = 5, what is the expected number of iterations
we need until it terminates? To estimate the solution produce 10,000 such random
walks and calculate the average number of iterations in the 10, 000 random walks you
produce. You'll want to turn the plot off here.

```{r}
n <- 10000
iterations <- rep(NA, n)
for (i in 1:n) {
  iterations[i] <- random.walk(plot.walk=FALSE)$num.steps
}
mean(iterations)
```

17. Modify your function random.walk() defined previously so that it takes an additional
argument seed: this is an integer that should be used to set the seed of the random number generator, before the random walk begins, with set.seed(). But, if seed is NULL, the default, then no seed should be set. Run your modified function
random.walk() function twice with the default inputs, then run it twice with the
input seed equal to (say) 33 and plot.walk = FALSE.

```{r}
random.walk <- function(x.start= 5, plot.walk = TRUE, seed=NULL){
  if (!is.null(seed)) {
    set.seed(seed)
  }
  x.vals <- 5
  i <- 1
  while (x.vals[i] > 0) {
    r <- runif(1, min = -2, max = 1)
    temp <- x.vals[i]
    if (temp + r < 0) {
      break
    }else {
      x.vals <- c(x.vals, x.vals[i] + r)
    }
    i <- i + 1
  }
  if (plot.walk) {
    plot(1:length(x.vals), x.vals, type = "o",
         xlab="Iteration", ylab="Random Walk")
  }
  return(list(x.vals = x.vals, num.steps = i))
}
random.walk()
random.walk()
random.walk(10, FALSE, seed =33)
random.walk(10, FALSE, seed=33)
```

Part 4: Monte Carlo Integration

Consider the integral:

$\int_0^1 g(x) dx = \int_0^1 e^{-x^{3}}dx$

18. Run the following code:

```{r}
g <- function(x) {
return(exp(-x^3))
}
x <- seq(0,1,.01)
alpha <- 2
beta <- 2
plot(x,g(x),type="l",xlab="x",ylab="",ylim=c(-.1,1.4))
polygon(c(0,seq(0,1,0.01),1),c(0,g(seq(0,1,0.01)),0) ,col="pink")
lines(x,rep(1,length(x)),col="red")
lines(x,dbeta(x,shape1=alpha,shape2=beta),col="blue")
legend("topleft",legend=c("g(x)","uniform","beta(2,2)"),
lty=c(1,1,1),col=c("black","red","blue"),cex=.6)

```

19. Using Monte Carlo Integration, approximate the integral $\int_0^1 e^{-x^{3}}dx$ using n =
$1000^2$ random draws from the distribution uniform(0,1).

```{r}
x <- runif(1000^2, 0, 1)
estimate <- mean(exp(-x^3))
estimate
```

20. Using Monte Carlo Integration, approximate the integral $\int_0^1 e^{-x^{3}}dx$ using n =
$1000^2$ random draws from the distribution beta($\alpha = 2$, $\beta = 2$).

```{r}
x <- rbeta(1000^2, 2, 2)
estimate <- mean(exp(-x^3))
estimate
```