---
title: "Lab 6"
author: "Ivan Wolansky, iaw2110"
date: "April 12, 2019"
output: pdf_document
---

# Instructions 
Make sure that you upload the PDF (or HTML) output after you have knitted the file. The files you upload to the Canvas page should be updated with commands you provide to answer each of the questions below.  You can edit this file directly to produce your final solutions.     

# Goal

The goal of this lab is to investigate the empirical behavior of a common hypothesis testing procedure through simulation using R. We consider the traditional two-sample t-test.  

# Two-Sample T-Test


Consider an experiment testing if a 35 year old male's heart rate statistically differs between a control group and a dosage group. Let $X$ denote the control group and let $Y$ denote the drug group.  One common method used to solve this problem is the two-sample t-test.  The null hypothesis for this study is:
$$H_0:\mu_1-\mu_2=\Delta_0,$$
where $\Delta_0$ is the hypothesized value.  The assumptions of the two sample pooled t-test follow below:  

## Assumptions 

\begin{enumerate}
\item $X_1,X_2,\ldots, X_m$ is a random sample from a normal distribution with mean $\mu_1$ and variance $\sigma^2_1.$
\item $Y_1,Y_2,\ldots, Y_n$ is a random sample from a normal distribution with mean $\mu_2$ and variance $\sigma^2_2.$
\item The $X$ and $Y$ samples are independent of one another.   
\end{enumerate}

##  Procedure 

The test statistic is 
$$t_{calc}=\frac{\bar{x}-\bar{y}-\Delta_0}{\sqrt{\frac{s^2_1}{m}+\frac{s^2_2}{n}}},$$
where $\bar{x},\bar{y}$ are the respective sample means and $s_1^2,s_2^2$ are the respective sample standard deviations. 

The approximate degrees of freedom is  
\[
df=\frac{\Big{(}\frac{s^2_1}{m}+\frac{s_2^2}{n}\Big{)}^2}{\frac{(s_1^2/m)^2}{m-1}+\frac{(s_2^2/n)^2}{n-1}}
\]
Under the null hypothesis, $t_{calc}$ has a student's t-distribution with $df$ degrees of freedom.     

##  Rejection rules 

\begin{table}[ht]
\begin{center}
\begin{tabular}{c|c}
Alternative Hypothesis   &  P-value calculation \\
\hline
&\\
$H_A: \mu_1-\mu_2>\Delta_0$  \ \ (upper-tailed) & $P(t_{calc}>T)$ \\
&\\
\hline
&\\
$H_A:\mu_1-\mu_2<\Delta_0$ \ \ (lower-tailed) & $P(t_{calc}<T)$ \\
&\\
\hline
&\\
$H_A: \mu_1-\mu_2 \neq\Delta_0$ \ \ (two-tailed) & $2*P(|t_{calc}|>T)$ \\
&\\
\hline
\end{tabular}
\end{center}
\end{table}

Reject $H_0$ when: $$Pvalue\leq \alpha$$ 

# Tasks

1) Using the **R** function **t.test**, run the two sample t-test on the following simulated dataset.  Note that the **t.test** function defaults a two-tailed alternative.  Also briefly interpret the output.    
```{r}
set.seed(5)
sigma=5
Control <- rnorm(30,mean=10,sd=sigma)
Dosage <- rnorm(35,mean=12,sd=sigma)
t.test(Control, Dosage)
```
1) (Interpretation) The fact that the sample means are smaller than the t-value means that the hypothesized mean should be against the null hypothesis and in favor of the alternative hypothesis.

2)  Write a function called **t.test.sim** that simulates **R** different samples of $X$ for control and **R** different samples of $Y$ for the drug group and computes the proportion of test statistics that fall in the rejection region.  The function should include the following:  
\begin{itemize}
\item Inputs:
\begin{itemize}
\item {\bf R} is the number of simulated data sets (simulated test statistics).  Let  {\bf R} have default 10,000.  
\item Parameters {\bf mu1}, {\bf mu2}, {\bf sigma1} and {\bf sigma2} which are the respective true means and true standard deviations of $X$ \& $Y$.  Let the parameters have respective defaults {\bf mu1=0}, {\bf mu2=0}, {\bf sigma1=1} and {\bf sigma2=1}.  
\item Sample sizes {\bf n} and {\bf m} defaulted at {\bf m=n=30}.  
\item \textbf{level} is the significance level as a decimal with default at $\alpha=.05$.  
\item \textbf{value} is the hypothesized value defaulted at 0.  
\end{itemize}
\item The output should be a \textbf{list} with the following labeled elements: 
\begin{itemize}  
\item \textbf{statistic.list} vector of simulated t-statistics (this should have length {\bf R}).  
\item \textbf{pvalue.list} vector of empirical p-values (this should have length {\bf R}).  
\item \textbf{empirical.size} is a single number that represents the proportion of simulated test statistics that fell in the rejection region.    
\end{itemize}
\end{itemize}

\pagebreak

I started the function below: 
```{r}
t.test.sim <- function(R=10000,
                           mu1=0,mu2=0,
                           sigma1=1,sigma2=1,
                           m=30,n=30,
                           level=.05,
                           value=0,
                           direction="Two") {
  
  #Define empty lists
  statistic.list <- rep(0,R)
  pvalue.list <- rep(0,R)
  
  for (i in 1:R) {
    
    # Sample realized data 
    Control <- rnorm(m, mean=mu1, sd=sigma1)
    Dosage <- rnorm(n, mean=mu2, sd=sigma2)
    
    # Testing values
    testing.procedure <- t.test(Control, Dosage, mu=value)
    statistic.list[i] <- testing.procedure$statistic
    pvalue.list[i] <- testing.procedure$p.value
   
  }
  
   size.list <- mean(pvalue.list <= level)
  return(list(statistic.list=statistic.list, pvalue.list=pvalue.list, empirical.size=size.list)) 
  
}
t.test.sim(R=10,mu1=10,mu2=12,sigma1=5, sigma2=5)
```
Evaluate your function with the following inputs 
**R=10**,**mu1=10**,**mu1=12**,**sigma1=5** and **sigma2=5**. 


3) Assuming the null hypothesis $$H_0:\mu_1-\mu_2=0$$ is true, compute the empirical size using 10,000 simulated data sets.  Use the function **t.test.sim** to accomplish this task and store the object as **sim**.  Output the empirical size quantity **sim$size**.  Comment on this value.  What is it close to?     

  **Note:**  use **mu1=mu1=10** (i.e., the null is true).  Also set **sigma1=5**,**sigma2=5** and **n=m=30**. 

```{r}
sim <- t.test.sim(R=10000, mu1=10, mu2=10, sigma1=5, sigma2=5, n=30, m=30)
sim$empirical.size
```
3) (Comment) This value of 0.0486 is very close to the level for $\alpha$ which is 0.05.

4) Plot a histogram of the simulated P-values, i.e., **hist(sim$pvalue.list)**.  What is the probability distribution shown from this histogram?  Does this surprise you? 

```{r}
hist(sim$pvalue.list)
```
4) (Solution) This does not surprise me because it shows that with the null hypothesis $H_{0} : \mu_{1} - \mu_{2} = 0$, the empirical density function of the p-values approaches the uniform distribution.

5) Plot a histogram illustrating the empirical sampling sampling of the t-statistic, i.e., **hist(sim$statistic.list,probability =TRUE)**.  What is the probability distribution shown from this histogram?    

```{r}
hist(sim$statistic.list,probability =TRUE)
```
5) (Solution) Here, the probability distribution shown approaches the standard normal distribution.

6) Run the following four lines of code:

      **t.test.sim(R=1000,mu1=10,mu2=10,sigma1=5,sigma2=5)$emperical.size**

      **t.test.sim(R=1000,mu1=10,mu2=12,sigma1=5,sigma2=5)$emperical.size**

      **t.test.sim(R=1000,mu1=10,mu2=14,sigma1=5,sigma2=5)$emperical.size**

      **t.test.sim(R=1000,mu1=10,mu2=16,sigma1=5,sigma2=5)$emperical.size**
      
      Comment on the results.  

```{r}
t.test.sim(R=1000,mu1=10,mu2=10,sigma1=5,sigma2=5)$empirical.size
t.test.sim(R=1000,mu1=10,mu2=12,sigma1=5,sigma2=5)$empirical.size
t.test.sim(R=1000,mu1=10,mu2=14,sigma1=5,sigma2=5)$empirical.size
t.test.sim(R=1000,mu1=10,mu2=16,sigma1=5,sigma2=5)$empirical.size

```
6) (Solution) When the sample sizes of m and n are fixed, the empirical probability of rejecting the null hypothesis increases as the data gets farther from the null hypothesis. This means that the empirical power of the test continues to increase as the effect size increases.

7) Run the following four lines of code:

      **t.test.sim(R=10000,mu1=10,mu2=12,sigma1=10,sigma2=10,m=10,n=10)$emperical.size**

      **t.test.sim(R=10000,mu1=10,mu2=12,sigma1=10,sigma2=10,m=30,n=30)$emperical.size**

      **t.test.sim(R=10000,mu1=10,mu2=12,sigma1=10,sigma2=10,m=50,n=50)$emperical.size**

      **t.test.sim(R=10000,mu1=10,mu2=12,sigma1=10,sigma2=10,m=100,n=100)$emperical.size**
      
      Comment on the results.  

```{r}
t.test.sim(R=10000,mu1=10,mu2=12,sigma1=10,sigma2=10,m=10,n=10)$empirical.size
t.test.sim(R=10000,mu1=10,mu2=12,sigma1=10,sigma2=10,m=30,n=30)$empirical.size
t.test.sim(R=10000,mu1=10,mu2=12,sigma1=10,sigma2=10,m=50,n=50)$empirical.size
t.test.sim(R=10000,mu1=10,mu2=12,sigma1=10,sigma2=10,m=100,n=100)$empirical.size

```
7) (Solution) This shows that when the difference in the means is fixed, the probability of rejecting the null hypothesis increases as the sample size increases. This means that if the effect size is fixed, then the empirical power of the test grows with the sample size.

8) **Extra credit:**  Modify the **t.test.sim()** function  to investigate how the power and size behave in the presence of heavy tailed data.  I.e., test how **robust** the t-test is in the presence of violations from normality.  **Hint:** the Cauchy distribution very heavy tailed.  

```{r}
t.test.sim.cauchy <- function(R=10000,
                           location1=0,location2=0,
                           scale1=1,scale2=1,
                           m=30,n=30,
                           level=.05,
                           value=0,
                           direction="Two") {
  
  #Define empty lists
  statistic.list <- rep(0,R)
  pvalue.list <- rep(0,R)
  
  for (i in 1:R) {
    
    # Sample realized data 
    Control <- rcauchy(m, location = location1, scale = scale1)
    Dosage <- rcauchy(n, location = location2, scale = scale2)
    
    # Testing values
    testing.procedure <- t.test(Control, Dosage, mu=value)
    statistic.list[i] <- testing.procedure$statistic
    pvalue.list[i] <- testing.procedure$p.value
   
  }
  
   size.list <- mean(pvalue.list <= level)
  return(list(statistic.list=statistic.list, pvalue.list=pvalue.list, empirical.size=size.list)) 
  
}
t.test.sim.cauchy(R=10,location1=10,location2=12,scale1=5, scale2=5)
```

```{r}
# varying location parameter
for (i in 1:15) {
  if (i == 1) {
  print(t.test.sim.cauchy(R=1000,location1=10,location2=10,scale1=5, scale2=5)$empirical.size)
  }
  else {
    print(t.test.sim.cauchy(R=1000,location1=10,location2=10 + i * 2,scale1=5, 
            scale2=5)$empirical.size)
  }
}

```
When the sample size is fixed, the empirical probability of rejecting the null increases as the location parameter departs from the null hypothesis. However, the t-test is clearly much less robust in this case, as location2 was varied from 10 to 40 and it still has relatively low empirical power with a heavy tailed distribution.

```{r}
#varying sample size
for (i in 1:15){
  if (i == 1) {
    print(t.test.sim.cauchy(R=10000,location1=10,location2=12,scale1=10, 
      scale2=10,m=10 ,n=10)$empirical.size)
  } else
    print(t.test.sim.cauchy(R=10000,location1=10,location2=12,scale1=10, 
        scale2=10 ,m=10 * (i * 10),n=10 * (i * 10))$empirical.size)
}
```
When the difference in the location parameter is fixed, the empirical probability of rejecting the null hypothesis varies between 0.02 and 0.25 when the sample size increases. This means that the sample size has no bearing on the probability of rejecting the null hypothesis. This shows the t-test is much less robust when increasing sample size with a heavy tailed distribution.